---
title: "Pitch Preference by Nationality"
format: html
execute:
  echo: false
  warning: false
  message: false
---

In this analysis, we explore whether MLB pitchers from different countries display noticeable preferences in pitch selection. Using Statcast pitch-level data from WBC rosters, we compare pitch usage rates across the top represented countries.

```{r}
library(tidyverse)
library(plotly)

# Load Statcast pitching data
pitch_data <- read_csv("data/statcast_pitchers.csv", show_col_types = FALSE)

# Map Statcast pitch abbreviations to full names
pitch_labels <- c(
  "FF" = "4-Seam Fastball",
  "SI" = "Sinker",
  "SL" = "Slider",
  "CH" = "Changeup",
  "FC" = "Cutter",
  "ST" = "Sweeper",
  "CU" = "Curveball",
  "FS" = "Splitter",
  "KC" = "Knuckle Curve",
  "SV" = "Slurve"
)

# Identify the top 6 countries by pitch volume
top_countries <- pitch_data |>
  count(country_name) |>
  slice_max(n, n = 6) |>
  pull(country_name)

# Calculate percentage usage for each pitch type by country
pitch_summary <- pitch_data |>
  filter(country_name %in% top_countries, pitch_type %in% names(pitch_labels)) |>
  mutate(pitch_name = pitch_labels[pitch_type]) |>
  count(country_name, pitch_name) |>
  group_by(country_name) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  ungroup()

# Build horizontal bar plot faceted across countries
p <- ggplot(pitch_summary, aes(
  x = pct,
  y = reorder(pitch_name, pct),
  fill = country_name,
  text = paste0(
    "<b>", country_name, "</b><br>",
    pitch_name, ": ", pct, "%<br>",
    "Pitches: ", scales::comma(n)
  )
)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~country_name, nrow = 2) +
  scale_x_continuous(labels = \(x) paste0(x, "%")) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Pitch Selection by Pitcher Nationality",
    subtitle = "Percentage of total pitches thrown by country (top 6 countries)",
    x = "Usage (%)",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  )

ggplotly(p, tooltip = "text")
```

## Pitch Characteristics by Country

Each point represents an individual pitch, showing release speed and spin rate across different pitch types for each top country.

```{r}
# Sample 400 pitches per country for a balanced and responsive plot
set.seed(42)
country_sample <- pitch_data |>
  filter(country_name %in% top_countries, pitch_type %in% names(pitch_labels)) |>
  drop_na(release_speed, release_spin_rate, player_name) |>
  slice_sample(n = 400, by = country_name) |>
  mutate(pitch_name = pitch_labels[pitch_type])

# Scatter plot of velocity vs spin rate faceted by country
p_scatter <- ggplot(country_sample, aes(
  x = release_speed,
  y = release_spin_rate,
  color = pitch_name,
  text = paste0(
    "<b>", player_name, "</b><br>",
    "Country: ", country_name, "<br>",
    "Pitch: ", pitch_name, "<br>",
    "Velocity: ", release_speed, " mph<br>",
    "Spin Rate: ", release_spin_rate, " rpm"
  )
)) +
  geom_point(alpha = 0.5, size = 1) +
  facet_wrap(~country_name, nrow = 2) +
  scale_color_brewer(palette = "Paired") +
  labs(
    title = "Pitch Velocity vs. Spin Rate by Country",
    subtitle = "Sample of individual pitches across the top 6 represented countries",
    x = "Release Speed (mph)",
    y = "Release Spin Rate (rpm)",
    color = "Pitch Type"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  )

ggplotly(p_scatter, tooltip = "text")
```


